% ── sections/00_frontmatter.tex ───────────────────────────────────────────────

% ── Title ─────────────────────────────────────────────────────────────────────
{\centering
  \fontsize{16}{19}\bfseries\selectfont
  Cancer Driver Gene Prediction via Explainable Multilayer Graph Neural Networks
  with Multi-Network Integration and Pathway Enrichment Analysis
  \par
}

\vspace{0.5em}

% ── Authors ───────────────────────────────────────────────────────────────────
{\centering
  \normalsize
  Yixi Wang\textsuperscript{1*}
  \par
}

\vspace{0.3em}

% ── Affiliations ──────────────────────────────────────────────────────────────
{\centering
  \small
  \textsuperscript{1}Department of Computer Science, School of Computing,
  National University of Singapore, 13 Computing Drive, Singapore 117417\\
  *Corresponding author: \url{c2h4wang@u.nus.edu}
  \par
}

\vspace{1em}

% ── Author Details Block ──────────────────────────────────────────────────────
\noindent\textit{Article Type:} Research Article

\noindent\textbf{Yixi Wang:}
Highest Qualification: Bachelor of Computing (in progress).
Department: Computer Science.
Post/Rank: Undergraduate student, Year 4, Computer Science.
Affiliation: School of Computing, National University of Singapore,
13 Computing Drive, Singapore 117417.
email: c2h4wang@u.nus.edu.

\vspace{1em}

% ── Abstract ──────────────────────────────────────────────────────────────────
\noindent\textbf{ABSTRACT}

\noindent
Identifying cancer driver genes from the vast landscape of somatic mutations is a central
challenge in computational oncology. We build upon and extend the Explainable Multilayer
Graph Neural Network (EMGNN), which fuses multi-omics node features with protein--protein
interaction (PPI) network topology via a two-stage graph neural network architecture.
Our work pursues five objectives. First, we reproduce the benchmark, achieving a mean
AUPR~=~0.743 (best~=~0.753) on the CPDB network (GCN backbone, 18 runs), confirming the
published trend. Second, we perform systematic ablation of four architectural
improvements---residual skip connections, Batch Normalisation, label smoothing
($\varepsilon$~=~0.05), and cosine learning-rate annealing---and find that BatchNorm1d is
detrimental in full-batch graph settings ($-$4.2\% AUPR); residual connections and label
smoothing are retained as beneficial components. Bayesian hyperparameter optimisation
(Optuna, 50 trials) identifies a superior configuration achieving AUPR~=~\textbf{0.8023}
(+5.4\% over baseline). Third, we extend the framework to all six PPI databases using
learnable per-network importance weights, achieving AUPR~=~\textbf{0.8067} and
AUROC~=~\textbf{0.9170}, a gain of \textbf{+5.9~percentage points} over the
single-network baseline. Learned network weights confirm CPDB and Multinet as the most
informative databases (weights~=~0.210 and 0.201). Fourth, we apply Integrated Gradients
attribution---revealing that DNA methylation and gene expression features dominate the
cancer signal---and Gene Set Enrichment Analysis via Enrichr, recovering 31 significant
Hallmark pathways (FDR~$<$~0.05); the highest-ranked terms include Epithelial Mesenchymal
Transition (FDR~=~$1.6\times10^{-17}$), PI3K/AKT/mTOR Signalling
(FDR~=~$5.8\times10^{-15}$), and Apoptosis (FDR~=~$6.4\times10^{-11}$). Fifth, we
implement eleven advanced GNN techniques---including Focal Loss for class imbalance,
heterophily-aware gated filtering, GraphMAE self-supervised pretraining, GPS Graph
Transformers for long-range meta-graph dependencies, cross-network attention fusion,
pathway hypergraph integration, and GNNExplainer for edge-level interpretability---as
modular, independently activatable components within the framework. The top predicted
genes---TP53, MUC16, CTNNB1, PIK3CA---are all well-established oncogenes or tumour
suppressors, validating the model's biological fidelity. These findings demonstrate that
multi-network integration, systematic hyperparameter optimisation, advanced GNN
architectures, and interpretability analysis are complementary strategies for building
robust, trustworthy cancer gene predictors.

\vspace{0.5em}
\noindent\textbf{Keywords:} cancer driver gene prediction, graph neural network,
protein--protein interaction network, multi-omics integration, gene set enrichment analysis,
model interpretability, hyperparameter optimisation, focal loss, graph transformer,
self-supervised pretraining

\vspace{1em}
