% ── sections/01_introduction.tex ─────────────────────────────────────────────

\section{Introduction}

Cancer arises through the accumulation of somatic mutations in a subset of genes---termed
\textit{cancer driver genes}---that confer selective growth advantages to cells. Distinguishing
these drivers from the large background of passenger mutations is essential for understanding
tumourigenesis, stratifying patients, and selecting targeted therapies~\cite{emogi2021}.
Genome-wide sequencing projects such as TCGA have catalogued thousands of mutations across
dozens of cancer types, yet the challenge of prioritising functionally relevant alterations
at scale remains largely unsolved~\cite{emgnn2023}.

Graph Neural Networks (GNNs) have emerged as a natural framework for this problem because
PPI networks provide a biologically meaningful relational structure: genes with similar
functions tend to cluster in network neighbourhoods, and disease genes are enriched in
specific modules~\cite{gcn2017}. Early GNN-based approaches such as EMOGI~\cite{emogi2021}
encode each gene as a node with multi-omics features and propagate information through a
single PPI network, achieving state-of-the-art precision--recall performance. The more
recent EMGNN framework~\cite{emgnn2023} generalises this to a multilayer setting by
constructing a \textit{meta-graph} whose nodes aggregate evidence from multiple per-network
GNN encoders, and further provides Integrated Gradients explanations at the edge and
feature level.

Despite these advances, several open questions remain. First, it is unclear which
architectural improvements---residual connections, normalisation, regularisation via label
smoothing, adaptive learning-rate schedules---translate to gains on biologically sparse,
full-batch graph tasks, where the entire network is processed in a single forward pass.
Second, while EMGNN nominally supports multiple PPI networks, the original evaluation is
primarily single-network; the extent of the multi-network gain and the role of per-network
importance weights have not been systematically quantified. Third, pathway-level
interpretability---connecting gene predictions to known cancer hallmarks---has not been fully
explored within the EMGNN framework~\cite{hallmarks2011}.

This work closes these gaps with a systematic study of the EMGNN framework that moves from
reproduction to improvement to interpretation. We first establish a faithful reproduction
of the original benchmark and then ask: which architectural choices actually help? Our
ablation and Bayesian hyperparameter search reveal that BatchNorm1d is \emph{harmful} in
full-batch graph settings ($-$4.2\% AUPR), while removing it together with tuned
hyperparameters yields AUPR~=~0.8023 (+5.4\% over baseline). We then show that training
across all six available PPI databases with learnable per-network importance weights
further improves performance to AUPR~=~0.8067 (+5.9\%), while the learned weights
quantify each database's relative contribution. Finally, Integrated Gradients and
GSEA analysis reveal that the model relies primarily on epigenetic (methylation) and
transcriptomic (gene expression) signals rather than mutation frequency, and that its
top-ranked genes are significantly enriched in canonical cancer hallmarks including EMT,
PI3K/AKT/mTOR, and G2-M checkpoint pathways.

