% ── sections/05_conclusions.tex ──────────────────────────────────────────────

\section{Conclusions}

This work presents a systematic investigation of cancer driver gene prediction using
multilayer graph neural networks, building upon and extending the EMGNN benchmark through
five interconnected methodological contributions.

Benchmark reproduction confirms that the GCN backbone achieves a mean AUPR of 0.743 (best
0.753) across 18 independent runs on CPDB, consistent with the published rank ordering of
GCN, GIN, and GAT backbones. The controlled ablation study of architectural improvements
yields an important negative finding: BatchNorm1d is harmful in the full-batch graph learning
setting and reduces AUPR by 4.2 percentage points; this result is reproducible and has
practical implications for practitioners applying normalisation layers to graph-structured
data where the entire graph is processed in a single forward pass. Residual skip connections
and label smoothing are confirmed as compatible improvements.

Bayesian hyperparameter optimisation (Optuna, 50 trials) identifies a configuration with
hidden dimension 32, four GNN layers, no BatchNorm, and a step learning-rate schedule,
achieving AUPR~=~\textbf{0.8023} on CPDB---surpassing the manual ablation and the
benchmark by 5.4 percentage points. This result demonstrates the importance of systematic
search over the architecture space, particularly for the hidden dimension and depth parameters.

The multi-network extension demonstrates that aggregating complementary PPI evidence
through a meta-graph provides a substantial and consistent performance improvement.
Training simultaneously on all six PPI databases with learned importance weights raises
AUPR from 0.748 to \textbf{0.8067} (+5.9\%), with AUROC reaching \textbf{0.9170}.
The learned per-network weights confirm the relative importance of each database: CPDB
and Multinet receive the highest weights (0.210 and 0.201), while IRefIndex (the sparsest
and oldest database) receives the lowest (0.096). This end-to-end differentiation of
network quality enables the model to adaptively exploit complementary evidence without
manual database pre-selection.

The interpretability analysis provides biological grounding for the model's predictions.
Integrated Gradients attribution reveals that DNA methylation and gene expression features
dominate the top-ranked importance scores, with METH:LIHC (hepatocellular carcinoma
methylation) and GE:BLCA (bladder cancer expression) as the two most discriminative
features. This finding suggests that epigenetic reprogramming and transcriptomic
dysregulation are the primary signals distinguishing cancer from non-cancer genes.
GSEA recovers 31 significant Hallmark pathways (FDR~$<$~0.05)---led by EMT
(FDR~=~$1.6\times10^{-17}$), PI3K/AKT/mTOR, and TGF-$\beta$---confirming that the
predicted cancer gene set is concentrated in established oncogenic programmes. The
high-confidence prediction of TP53, MUC16, CTNNB1, PIK3CA, and BRCA1 at the top of
the list validates biological fidelity.

Beyond the core four-methodology pipeline, we implement eleven advanced GNN techniques
(Methodology~5) as modular, independently activatable extensions. These address key
limitations of the baseline: Focal Loss and label smoothing tackle the 1:30 class
imbalance; DropEdge and heterophily-aware gated filtering address over-smoothing in the
low-homophily PPI setting; GraphMAE pretraining exploits unlabelled graph structure for
better weight initialisation; GPS Graph Transformers and cross-network attention enable
long-range and gene-specific multi-network reasoning; pathway hypergraph encoding and
PINNACLE protein embeddings incorporate external biological knowledge; and GNNExplainer
provides edge-level interpretability complementing the feature-level Integrated Gradients.
These components are designed for compositional experimentation: any subset can be activated
simultaneously via command-line flags without code changes.

Several limitations should be noted. The Methodology~5 techniques are implemented and
unit-tested but not yet evaluated in a systematic ablation on the six-network configuration
due to computational constraints; this is the most immediate next step.
Results are reported without averaging over multiple random seeds for the multi-network
and Optuna experiments, which introduces run-to-run variability. Three external-data-dependent
modules (pathway hypergraph, HIPGNN eigenvectors, PINNACLE embeddings) require downloading
and preprocessing external datasets before they can be activated.

Taken together, these results demonstrate that: (i) multi-network integration consistently
outperforms single-network baselines regardless of model variant; (ii) appropriate
hyperparameter search and architectural choices (no BatchNorm, residual connections, label
smoothing) amplify the gains further; (iii) the resulting predictions are biologically
interpretable and pathway-coherent; and (iv) the framework is extensible to frontier GNN
techniques through a modular, flag-driven architecture. The code, exact package versions,
trained model artefacts, and reproducible experiment configurations are released publicly
to support further development of interpretable cancer gene prediction models.
