% ── sections/07_references.tex ───────────────────────────────────────────────
% Reference list: Times New Roman, 8pt, Justified, IEEE-with-URL style.
% Citations in text use \cite{refN} matching the item numbers below.

\section*{References}

{\fontsize{8}{10}\selectfont
\begin{enumerate}

  \item M. Chatzianastasis, M. Vazirgiannis, and Z. Zhang,
        ``Explainable Multilayer Graph Neural Network for Cancer Gene Prediction,''
        \textit{Bioinformatics}, vol.~39, no.~11, p.~btad643, Nov. 2023.
        \url{https://doi.org/10.1093/bioinformatics/btad643}  % cite: emgnn2023

  \item R. Schulte-Sasse, S. Budach, D. Hnisz, and A. Marsico,
        ``Integration of multiomics data with graph convolutional networks to identify
        new cancer genes and their associated molecular mechanisms,''
        \textit{Nat. Mach. Intell.}, vol.~3, no.~6, pp.~513--526, Jun. 2021.
        \url{https://doi.org/10.1038/s42256-021-00325-y}  % cite: emogi2021

  \item T. N. Kipf and M. Welling,
        ``Semi-Supervised Classification with Graph Convolutional Networks,''
        in \textit{Proc. Int. Conf. Learn. Represent. (ICLR)}, 2017.
        \url{https://arxiv.org/abs/1609.02907}  % cite: gcn2017

  \item P. Velickovic, G. Cucurull, A. Casanova, A. Romero, P. Lio, and Y. Bengio,
        ``Graph Attention Networks,''
        in \textit{Proc. Int. Conf. Learn. Represent. (ICLR)}, 2018.
        \url{https://arxiv.org/abs/1710.10903}  % cite: gat2018

  \item K. Xu, W. Hu, J. Leskovec, and S. Jegelka,
        ``How Powerful are Graph Neural Networks?''
        in \textit{Proc. Int. Conf. Learn. Represent. (ICLR)}, 2019.
        \url{https://arxiv.org/abs/1810.00826}  % cite: gin2019

  \item W. L. Hamilton, Z. Ying, and J. Leskovec,
        ``Inductive Representation Learning on Large Graphs,''
        in \textit{Advances in Neural Information Processing Systems (NeurIPS)},
        vol.~30, 2017.
        \url{https://arxiv.org/abs/1706.02216}  % cite: sage2017

  \item M. Sundararajan, A. Taly, and Q. Yan,
        ``Axiomatic Attribution for Deep Networks,''
        in \textit{Proc. Int. Conf. Mach. Learn. (ICML)}, pp.~3319--3328, 2017.
        \url{https://arxiv.org/abs/1703.01365}  % cite: ig2017

  \item A. Subramanian \textit{et al.},
        ``Gene Set Enrichment Analysis: A Knowledge-Based Approach for Interpreting
        Genome-Wide Expression Profiles,''
        \textit{Proc. Natl. Acad. Sci. U.S.A.}, vol.~102, no.~43, pp.~15545--15550,
        Oct. 2005.
        \url{https://doi.org/10.1073/pnas.0506580102}  % cite: gsea2005

  \item E. Y. Chen \textit{et al.},
        ``Enrichr: Interactive and Collaborative HTML5 Gene List Enrichment Analysis Tool,''
        \textit{BMC Bioinformatics}, vol.~14, p.~128, Apr. 2013.
        \url{https://doi.org/10.1186/1471-2105-14-128}  % cite: enrichr2013

  \item A. Liberzon \textit{et al.},
        ``The Molecular Signatures Database Hallmark Gene Set Collection,''
        \textit{Cell Syst.}, vol.~1, no.~6, pp.~417--425, Dec. 2015.
        \url{https://doi.org/10.1016/j.cels.2015.12.004}  % cite: msigdb2015

  \item D. Hanahan and R. A. Weinberg,
        ``Hallmarks of Cancer: The Next Generation,''
        \textit{Cell}, vol.~144, no.~5, pp.~646--674, Mar. 2011.
        \url{https://doi.org/10.1016/j.cell.2011.02.013}  % cite: hallmarks2011

  \item D. P. Kingma and J. Ba,
        ``Adam: A Method for Stochastic Optimization,''
        in \textit{Proc. Int. Conf. Learn. Represent. (ICLR)}, 2015.
        \url{https://arxiv.org/abs/1412.6980}  % cite: kingma2014adam

  \item S. Ioffe and C. Szegedy,
        ``Batch Normalization: Accelerating Deep Network Training by Reducing
        Internal Covariate Shift,''
        in \textit{Proc. Int. Conf. Mach. Learn. (ICML)}, pp.~448--456, 2015.
        \url{https://arxiv.org/abs/1502.03167}  % cite: batchnorm2015

  \item K. He, X. Zhang, S. Ren, and J. Sun,
        ``Deep Residual Learning for Image Recognition,''
        in \textit{Proc. IEEE Conf. Comput. Vis. Pattern Recognit. (CVPR)},
        pp.~770--778, 2016.
        \url{https://doi.org/10.1109/CVPR.2016.90}  % cite: resnet2016

  \item I. Loshchilov and F. Hutter,
        ``SGDR: Stochastic Gradient Descent with Warm Restarts,''
        in \textit{Proc. Int. Conf. Learn. Represent. (ICLR)}, 2017.
        \url{https://arxiv.org/abs/1608.03983}  % cite: cosine2017

  \item T. Cai, S. Luo, K. Xu, D. He, T.-Y. Liu, and L. Wang,
        ``GraphNorm: A Principled Approach to Accelerating Graph Neural Network Training,''
        in \textit{Proc. Int. Conf. Mach. Learn. (ICML)}, pp.~1204--1215, 2021.
        \url{https://proceedings.mlr.press/v139/cai21e.html}  % cite: graphnorm2021

  \item S. A. Forbes \textit{et al.},
        ``COSMIC: Somatic Cancer Genetics at High-Resolution,''
        \textit{Nucleic Acids Res.}, vol.~45, no.~D1, pp.~D777--D783, Jan. 2017.
        \url{https://doi.org/10.1093/nar/gkw1121}  % cite: cgc2022

  \item N. Kokhlikyan \textit{et al.},
        ``Captum: A Unified and Generic Model Interpretability Library for PyTorch,''
        \textit{arXiv preprint arXiv:2009.07896}, 2020.
        \url{https://arxiv.org/abs/2009.07896}  % cite: captum2020

  \item M. S. Lawrence \textit{et al.},
        ``Mutational heterogeneity in cancer and the search for new cancer-associated genes,''
        \textit{Nature}, vol.~499, no.~7457, pp.~214--218, Jul. 2013.
        \url{https://doi.org/10.1038/nature12213}  % cite: lawrence2013

  \item T.-Y. Lin, P. Goyal, R. Girshick, K. He, and P. Doll\'{a}r,
        ``Focal Loss for Dense Object Detection,''
        in \textit{Proc. IEEE Int. Conf. Comput. Vis. (ICCV)}, pp.~2980--2988, 2017.
        \url{https://arxiv.org/abs/1708.02002}  % cite: focal2017

  \item Y. Rong, W. Huang, T. Xu, and J. Huang,
        ``DropEdge: Towards Deep Graph Convolutional Networks on Node Classification,''
        in \textit{Proc. Int. Conf. Learn. Represent. (ICLR)}, 2020.
        \url{https://arxiv.org/abs/1907.10903}  % cite: dropedge2020

  \item Z. Hou \textit{et al.},
        ``GraphMAE: Self-Supervised Masked Graph Autoencoders,''
        in \textit{Proc. ACM SIGKDD Int. Conf. Knowl. Discov. Data Min.},
        pp.~594--604, 2022.
        \url{https://arxiv.org/abs/2205.10803}  % cite: graphmae2022

  \item L. Ramp\'{a}\v{s}ek \textit{et al.},
        ``Recipe for a General, Powerful, Scalable Graph Transformer,''
        in \textit{Advances in Neural Information Processing Systems (NeurIPS)},
        vol.~35, pp.~14501--14515, 2022.
        \url{https://arxiv.org/abs/2205.12454}  % cite: gps2022

  \item Y. Feng, H. You, Z. Zhang, R. Ji, and Y. Gao,
        ``Hypergraph Neural Networks,''
        in \textit{Proc. AAAI Conf. Artif. Intell.}, vol.~33, pp.~3558--3565, 2019.
        \url{https://doi.org/10.1609/aaai.v33i01.33013558}  % cite: hgnn2019

  \item M. Li \textit{et al.},
        ``Context-aware protein structure and function prediction with PINNACLE,''
        \textit{Nat. Methods}, vol.~21, pp.~766--776, 2024.
        \url{https://doi.org/10.1038/s41592-024-02186-6}  % cite: pinnacle2024

  \item Z. Ying, D. Bourgeois, J. You, M. Zitnik, and J. Leskovec,
        ``GNNExplainer: Generating Explanations for Graph Neural Networks,''
        in \textit{Advances in Neural Information Processing Systems (NeurIPS)},
        vol.~32, 2019.
        \url{https://arxiv.org/abs/1903.03894}  % cite: gnnexplainer2019

  \item M. Dwivedi \textit{et al.},
        ``Graph Neural Networks with Learnable Structural and Positional Representations,''
        in \textit{Proc. Int. Conf. Learn. Represent. (ICLR)}, 2022.
        \url{https://arxiv.org/abs/2110.07875}  % cite: rwpe2020

  \item Z. Zhou, Y. Liang, and H. Huang,
        ``HIPGNN: High-frequency Information Preserved Graph Neural Networks
        for Graph Anomaly Detection,''
        in \textit{Proc. AAAI Conf. Artif. Intell.}, 2025.
        % cite: hipgnn2025

\end{enumerate}
}
